% -*-latex-*-
\chapter{Program: \io{ci2ptx}\label{ch.ci2ptx}}

\section{Purpose} Plot a confidence region using \PiCTeX\ graphics

\section{Usage\label{sec.ci2ptx.usage}} \io{ci2ptx [options] [inputfile]}\\
where options may include:
\begin{description}
\item[\io{-h$<$number$>$}] Set plot height to \verb@<number>@ inches.
\item[\io{-w$<$number$>$}] Set plot width to \verb@<number>@ inches.
\end{description}
By default, the program reads standard input.

\section{Description}

\io{ci2ptx} reads an input file describing a confidence region, which
must be in the format of the output produced by \io{mmci}, and writes
a series of \PiCTeX{} commands to the standard output.  These commands
generate a graphical display of the confidence region when processed
by the text-formatting program \LaTeX\ \cite{Lamport:LDP-94} together with the
\PiCTeX{} macros \cite{Wichura:PM-87}.  These are both available from
the address listed in the bibliography.

To produce a graphical display, proceed as follows:
\begin{enumerate}
\item Execute the command 
\begin{verbatim}
  ci2ptx csw.ci > figxxx.tex
\end{verbatim}
to place \PiCTeX{} code into the file \io{figxxx.tex}.  

\item Copy the file mkfig.tex from the directory that contains the
    mismatch source code into your current directory.  \io{mkfig.tex} is
    a very brief file that tells \LaTeX{} to read in the file
    \io{figxxx.tex}, which you just created.

\item Make sure that your current directory contains a file called
      \io{mkfig.tex} containing the following:
\begin{verbatim}
% -*-latex-*-
\documentclass[]{article}
\usepackage{pictex}
\begin{document}
\input{figxxx}       %% Edit this line to match your filename
\end{document}
\end{verbatim}
If you change the name of \io{figxxx.tex}, you must edit
\io{mkfig.tex} to reflect this change.

\item Type \io{latex mkfig} to run \LaTeX.  This produces a file called
    \io{mkfig.dvi}.

\item To look at the resulting plot on your screen, try typing
      \io{xdvi~mkfig} if you are working under X-windows.  If this
      doesn't work, or if you are working in some other environment,
      ask your local computer guru how to preview the ``.dvi'' files
      that are produced by \TeX.

\item On my system, the next step is to type \io{dvips~mkfig}.  This
is really two steps rolled into one.  First, \io{dvips} translates the
.dvi format into the format that my printer understands.  Second, it
passes the translation to the printer.  This may require two steps on
your system.
\end{enumerate}

\subsection{Customizing the figure}

If you don't like the way the figure looks, there are two things you
can do.

\subsubsection{Customizing with command line arguments}
 
To make the individual plots 0.75 by 0.75 inches, use the following
command:
\begin{verbatim}
  ci2ptx -w0.75 -h0.75 csw.ci > figxxx.tex
\end{verbatim}

\subsubsection{Editing the \PiCTeX{} file}

You can also change the way the figure looks by editing the \PiCTeX{}
code in \io{figxxx.tex}.  For example, the lines that read
\begin{verbatim}
%% Use the following to hide the 1% region
%\def\99pct{\reject}
\end{verbatim}
provide a means of making the 99 percent confidence region show on the
graph.  All you have to do is to remove the ``\%'' from the second of
these lines.  The lines that read
\begin{verbatim}
%% Definitions:
\def\accept{{$\bullet$}}
\def\99pct{{$\cdot$}}
\def\reject{{\footnotesize$\circ$}}
\end{verbatim}
provide a means of changing the symbols that are used to represent
points inside and outside of the confidence region.  Change these
definitions to whatever suits you.


