% -*-latex-*-
\chapter{Program: \io{ci2txt}\label{ch.ci2txt}}

\section{Purpose} Plot a confidence region using character graphics

\section{Usage\label{sec.ci2txt.usage}} \io{ci2txt [inputfile]}\\
By default, the program reads standard input.

\section{Description}

\io{ci2txt} reads an input file describing a confidence region, which
must be in the format of the output produced by \io{mmci}, and draws a
graph of the confidence region on standard output.  The graph is
produced using standard text characters.

The output consistes of several panels, one for each level of the
growth parameter.  For example, if the input to \io{mmci} contained
the line
\begin{verbatim}
RangeGrowth = 0 3 1 ;
\end{verbatim}
then the four values of the growth parameter will be considered by
\io{mmci}: 0, 1, 2, and 3.  Accordingly, there will be four panels in
the output of \io{ci2txt}.  The first panel look like this:
\begin{alltt}
\begin{minipage}{\textwidth}
Definitions:
 O : Reject at 1 percent significance level
 . : Within 99 percent confidence region
 * : Within 95 percent confidence region

Panel 0: N1/N0 = 10^0:

l          +-----------+
o    3.00  |      O    |
g    2.50  |      O    |
     2.00  |      O    |
t    1.50  |      O    |
h    1.00  |      *    |
e    0.50  |      *    |
t    0.00  |      O    |
a          +-----------+
0                       
              undefined 
\end{minipage}
\end{alltt}
This panel summarizes simulations for the case in which no growth
occurred, so that $\THETAa = \THETAb$.  The vertical axis shows values
of $\CLog \THETAa$.  The horizontal axis is usually for $\TAU$, but it
is undefined here because this panel is for the case in which no
growth has occurred.  Within the plot, each symbol indicates the
result of a hypothesis test.  Open circles ``O'' indicate hypotheses
that have been rejected at the 0.01 significance level, dots ``.''
indicate hypotheses that have been rejected at the 0.05 significance
level, and asterisks ``*'' indicate hypotheses that have not been
rejected.   The plot shown here indicates that we cannot reject the
hypotheses that $\CLog \THETA = 0.5$ or that $\CLog \THETA = 1$.  All
other hypotheses involving a constant population size are rejected.

The second panel in the output of \io{ci2txt} looks like
\begin{alltt}
\begin{minipage}{\textwidth}
Panel 1: N1/N0 = 10^1:

l          +-----------+
o    3.00  |O O O O O O|
g    2.50  |O O O O O O|
     2.00  |O O O O O O|
t    1.50  |O O O O O O|
h    1.00  |* * * . O O|
e    0.50  |* * * * * O|
t    0.00  |O * * * * *|
a          +-----------+
0                   1 1 
            2 4 6 8 0 2 
                        
                    tau 
\end{minipage}
\end{alltt}

